---
title: "ME315_Day01"
output: html_document
---


## Basic R Commands

The commands below

1. create vectors x and y

2. Returning the gtheir length

3. sum them together

4. checking the objects present in the environment

5. removing x and y from the environment

6. create a 2X2 matrix with numbers from 1 to 4

7. find the square root and the square of each matrix's component

8. Drawing random numbers from the Normal distribution

9. Compute summary statistics such as sample mean, variance, standard deviation and correlation

```{r setup, include=TRUE}
x <- c(1,3,2,5)
x
y = c(1,4,3)
y
length(x)
length(y)
x+y
ls()
rm(x,y)
ls()
rm(list=ls())
?matrix
x=matrix(data=c(1,2,3,4), nrow=2, ncol=2)
x
x=matrix(c(1,2,3,4),2,2)
matrix(c(1,2,3,4),2,2,byrow=TRUE)
sqrt(x)
x^2
x=rnorm(50)
y=x+rnorm(50,mean=50,sd=.1)
cor(x,y)
set.seed(1303)
rnorm(50)
set.seed(3)
y=rnorm(100)
mean(y)
var(y)
sqrt(var(y))
sd(y)
```

## Activity 1

Draw 10 numbers from the normal distribution with mean 5 and variance 2. Store them in the vector x and calculate the sample mean and standard deviation. Put them in a 2 X 5 matrix by row and and raise each number to the power of three.

## Graphics

The commands below show how to:
1. Do a scatter plot with the command `plot` and save it to a pdf file.
2. Create sequences and store them into vectors
3. Do a contour plot with the command `contour`. This requires three arguments: a vector of the x values, the y values and a matrix whose values correspond to each (x,y) pair (coordinates). Also do a heatmap with the command `image`.
4. Produce a 3-d plot for the above with the commande `persp`. The two arguments theta and phi control the angles at which the plot is viewed. 

```{r}
x=rnorm(100)
y=rnorm(100)
plot(x,y)
plot(x,y,xlab="this is the x-axis",ylab="this is the y-axis",main="Plot of X vs Y")
pdf("Figure.pdf")
plot(x,y,col="green")
dev.off()
x=seq(1,10)
x
x=1:10
x
x=seq(-pi,pi,length=50)
y=x
f=outer(x,y,function(x,y)cos(y)/(1+x^2))
contour(x,y,f)
contour(x,y,f,nlevels=45,add=T)
fa=(f-t(f))/2
contour(x,y,fa,nlevels=15)
image(x,y,fa)
persp(x,y,fa)
persp(x,y,fa,theta=30)
persp(x,y,fa,theta=30,phi=20)
persp(x,y,fa,theta=30,phi=70)
persp(x,y,fa,theta=30,phi=40)
```


### Activity 2 

Construct x and y to be sequences of length 100 from -5 to 5. Then
1. Set f1 to be 2 times x and do a scatter plot of x and f1.
2. Set f2 to be 2 times x minus 3 times y and produce a contour plot, a heatmap and 3d plot of f2 vs x and y


```{r, include=FALSE}
x=seq(-5,5,length=100)
y=x
f1=2*x
plot(x,f1)
f2=outer(x,y,function(x,y)2*x-3*y)
plot(x,f1)
contour(x,y,f2)
contour(x,y,f2,nlevels=45,add=T)
image(x,y,f2)
persp(x,y,f2)
```
## Indexing Data

Below are some commands for practicing in working with matrices and accessing their elements

```{r}
A=matrix(1:16,4,4)
A
A[2,3]
A[c(1,3),c(2,4)]
A[1:3,2:4]
A[1:2,]
A[,1:2]
A[1,]
A[-c(1,3),]
A[-c(1,3),-c(1,3,4)]
dim(A)
```

## Loading Data

Below are some commands for loading data from an existing file.

```{r}
Auto=read.table("Auto.data.txt")
dim(Auto)
names(Auto)
Auto=read.table("Auto.data.txt",header=T,na.strings="?")
dim(Auto)
names(Auto)
Auto=read.csv("Auto.csv",header=T,na.strings="?")
dim(Auto)
names(Auto)
Auto[1:4,]
Auto=na.omit(Auto)
dim(Auto)
names(Auto)
```

## Additional Graphical and Numerical Summaries

Below are some commands to do exploratory analysis on the Auto dataset

```{r}
Auto=read.csv("Auto.csv",header=T,na.strings="?")
plot(Auto$cylinders, Auto$mpg)
attach(Auto)
plot(cylinders, mpg)
cylinders=as.factor(cylinders)
plot(cylinders, mpg)
plot(cylinders, mpg, col="red")
plot(cylinders, mpg, col="red", varwidth=T)
plot(cylinders, mpg, col="red", varwidth=T,horizontal=T)
plot(cylinders, mpg, col="red", varwidth=T, xlab="cylinders", ylab="MPG")
hist(mpg)
hist(mpg,col=2)
hist(mpg,col=2,breaks=15)
pairs(Auto)
pairs(~ mpg + displacement + horsepower + weight + acceleration, Auto)
plot(horsepower,mpg)
identify(horsepower,mpg,name)
summary(Auto)
summary(mpg)
```

## Activity 3

This exercise involves the Boston housing data set which is
part of the MASS library in R. You can read about it typing `?Boston` after loading the library. To load and view part of it use the following commands:

```{r}
library(MASS)
head(Boston)
```


(a) How many rows are in this data set? How many columns? What do the rows and columns represent?

(b) Make some pairwise scatterplots of the predictors (columns) in this data set. Describe your findings.

(c) Are any of the predictors associated with per capita crime rate? If so, explain the relationship.

(d) Do any of the suburbs of Boston appear to have particularly high crime rates? Tax rates? Pupil-teacher ratios? Comment on the range of each predictor.

(e) How many of the suburbs in this data set bound the Charles river?

(f) What is the median pupil-teacher ratio among the towns in this data set?

(g) Which suburb of Boston has lowest median value of owner- occupied homes? What are the values of the other predictors for that suburb, and how do those values compare to the overall ranges for those predictors? Comment on your findings.

(h) In this data set, how many of the suburbs average more than seven rooms per dwelling? More than eight rooms per dwelling? Comment on the suburbs that average more than eight rooms per dwelling.


```{r, include=FALSE}
#(a)
library(MASS)
?Boston
dim(Boston)
# 506 rows, 14 columns
# 14 features, 506 housing values in Boston suburbs

# (b)
pairs(Boston)
# X correlates with: a, b, c
# crim: age, dis, rad, tax, ptratio
# zn: indus, nox, age, lstat
# indus: age, dis
# nox: age, dis
# dis: lstat
# lstat: medv

# (c)
plot(Boston$age, Boston$crim)
# Older homes, more crime
plot(Boston$dis, Boston$crim)
# Closer to work-area, more crime
plot(Boston$rad, Boston$crim)
# Higher index of accessibility to radial highways, more crime
plot(Boston$tax, Boston$crim)
# Higher tax rate, more crime
plot(Boston$ptratio, Boston$crim)
# Higher pupil:teacher ratio, more crime

# (d)
par(mfrow=c(1,3))
hist(Boston$crim[Boston$crim>1], breaks=25)
# most cities have low crime rates, but there is a long tail: 18 suburbs appear
# to have a crime rate > 20, reaching to above 80
hist(Boston$tax, breaks=25)
# there is a large divide between suburbs with low tax rates and a peak at 660-680
hist(Boston$ptratio, breaks=25)
# a skew towards high ratios, but no particularly high ratios

# (e)
dim(subset(Boston, chas == 1))


# (f)
median(Boston$ptratio)


# (g)
t(subset(Boston, medv == min(Boston$medv)))

#              399      406
# crim     38.3518  67.9208 above 3rd quartile
# zn        0.0000   0.0000 at min
# indus    18.1000  18.1000 at 3rd quartile
# chas      0.0000   0.0000 not bounded by river
# nox       0.6930   0.6930 above 3rd quartile
# rm        5.4530   5.6830 below 1st quartile
# age     100.0000 100.0000 at max
# dis       1.4896   1.4254 below 1st quartile
# rad      24.0000  24.0000 at max
# tax     666.0000 666.0000 at 3rd quartile
# ptratio  20.2000  20.2000 at 3rd quartile
# black   396.9000 384.9700 at max; above 1st quartile
# lstat    30.5900  22.9800 above 3rd quartile
# medv      5.0000   5.0000 at min
summary(Boston)
# Not the best place to live, but certainly not the worst.

# (h)
dim(subset(Boston, rm > 7))

dim(subset(Boston, rm > 8))

summary(subset(Boston, rm > 8))
summary(Boston)

# relatively lower crime (comparing range), lower lstat (comparing range)
```
