---
title: "ME315-day01"
output: html_document
---
      
In this workshop:

- Basic commands for vectors and matrices
- Data visualisation
- Loading data and EDA

## Basic R Commands

**Vectors and sequences**

- Create vectors x and y
- Vector arithmetic


```{r setup, include=TRUE}

x <- c(1,3,2,5)
y <-  c(1,4,3,0)

length(x)
length(y)

x+y
2*x-y+1
```

The function `seq` creates a sequence of numbers

```{r}
#sequences
x <- seq(1,10)

#sequence with specified number of values
y <- seq(1,10,length.out=20)

```

`rm` is the function that removes objects from the environment

```{r}
rm(x,y)  #to remove objects x and y
#rm(list=ls())
```


**Matrices**

The function `matrix` creates a matrix object. 

To construct a 2X2 matrix with numbers from 1 to 4, do:


```{r}
?matrix
x=matrix(data=c(1,2,3,4), nrow=2, ncol=2)
x
x=matrix(c(1,2,3,4),2,2)

```

**Elementwise operations on matrices**

To find the square root and the square of each matrix's component:

```{r}
# Elementwise operations on matrices
sqrt(x)
x^2
```

****


## Indexing and slicing 

Below are some commands for practicing in working with matrices and accessing their elements

```{r}
A=matrix(1:16,4,4)
A
A[2,3] 
A[c(1,3),c(2,4)]
A[1:3,2:4]
A[1:2,]
A[,1:2]
A[1,]
A[-c(1,3),]
A[-c(1,3),-c(1,3,4)]
dim(A)
```


## Drawing observations from the normal distribution

```{r}

x=rnorm(50)

#for replicability set seed
set.seed(1303)
rnorm(50)

set.seed(3)
y=rnorm(100)

median(y)
mean(y)

var(y)
sqrt(var(y))
sd(y)

min(x)
max(x)

```

**Correlations**

Recall the Pearson correlation measures the strength of linear relationshiop between 2 quantitative variables, and takes values [-1,1].

Task: Create 2 variables: 50 observations from $X \sim N(0,1)$, and $Y=X+N(50,0.1^2)$
Then compute the (Pearson's) correlation.

```{r}

y=x+rnorm(50,mean=50,sd=.1)
cor(x,y)
```


## Activity 1

1. Run the following chunk:

```{r}
A <- matrix(c(1,2,3,4),2,2,byrow=TRUE)
t(A) 
```

What does the function `t` do?

### Add your reply here:

2. Draw 10 numbers from the normal distribution with mean 5 and variance 2. Store them in the vector x and calculate the sample mean and standard deviation. Put them in a 2 X 5 matrix by row and and raise each number to the power of three.

###  Add your code here:


## Data visualisation

R has a set of functions for creating plots. A popular package for data visualisation is ggplot. We will focus on the core R graphics.


3. Do a contour plot with the command `contour`. This requires three arguments: a vector of the x values, the y values and a matrix whose values correspond to each (x,y) pair (coordinates). Also do a heatmap with the command `image`.
4. Produce a 3-d plot for the above with the commande `persp`. The two arguments theta and phi control the angles at which the plot is viewed. 

**Scatterplot**

1st simulate data for 2 variables.

```{r}
x1 <- rnorm(100)
y1 <- rnorm(100)

```

`plot` produces a scatterplot

```{r}
plot(x1,y1) #note order matters

```

To save the plot as a pdf in your folder, do:

```{r}
pdf("Figure.pdf")
plot(x,y,col="blue")
dev.off()  #closes the plot and uploads it 
```


## Loading Data

The functions for loading data depends on the format of the data.

- `read.table` is used to load data that is stored as a txt or csv file. 
- `read.csv` is for csv

A common error when loading data is not checking whether or not the data comes with a header. 
Also, if you had not already done so at the beginning of your session, set the working directory to the folder where your files are stored.

**Example: Automobile data**

The data Auto.data.txt consists of variables on the profile of vehicles. The data set contains a header (check). 
Data sets may have variables with missing values. Here, `na.strings=?` sets anything with ? to NA.

Load the data and check all us ok using `head`:

```{r}
Auto <- read.table("Auto.data.txt", header=T,na.strings="?")
head(Auto)
```

`summary` gives the 5-point summary statistics  plus the mean for each variable.

```{r}
summary(Auto)
```

**Exploratory data analysis (EDA)** is the first to fitting a model. EDA is an iterative cycle. You:

- Generate questions about your data.

- Search for answers by visualising, transforming, and modelling your data.

- Use what you learn to refine your questions and/or generate new questions.

Plots are important in EDA. The suitability of plots depend on the variable type.

**univariate plots**
Continuous variables: histogram, QQ-plot
Factors: bar charts, pie-charts

**bivariate plots**
Continuous-continuous: scatterplot
Continuous-factor: boxplot

Examples:

mpg is continuous. A histogram for mpg is

```{r}
hist(Auto$mpg)
```

Note the use of $. You may think of "Auto$mpg" as the attribute called mpg contained in object Auto.

```{r}
pairs(Auto)

```

```{r}
pairs(~ mpg + displacement + horsepower + weight + acceleration, Auto)

```


**Subsetting data**

How many vehicles have mpg>30? 
What is the distribution of mpg with horsepower = 90?

We can answer such question by using functions for subsetting the data. There is more than one way to subset data and it's a matter of preference.

The number of vehicles is:

```{r}
nrow(subset(Auto,mpg>10)) #or use dim
```

The histogram for mpg with horsepower =100 is:

```{r}
hist(Auto$mpg[Auto$horsepower==90]) # note the double =
```


## Activity 2

This exercise involves the Boston housing data set which is
part of the MASS library in R. You can read about it typing `?Boston` after loading the library. To load and view part of it use the following commands:

```{r}
library(MASS)
head(Boston)
```


(a) How many rows are in this data set? How many columns? What do the rows and columns represent?

(b) Make some pairwise scatterplots of the predictors (columns) in this data set. Describe your findings.

(c) Are any of the predictors associated with per capita crime rate? If so, explain the relationship.

(d) Do any of the suburbs of Boston appear to have particularly high crime rates? Tax rates? Pupil-teacher ratios? Comment on the range of each predictor.

(e) How many of the suburbs in this data set bound the Charles river?

(f) What is the median pupil-teacher ratio among the towns in this data set?

(g) Which suburb of Boston has lowest median value of owner- occupied homes? What are the values of the other predictors for that suburb, and how do those values compare to the overall ranges for those predictors? Comment on your findings.

(h) In this data set, how many of the suburbs average more than seven rooms per dwelling? More than eight rooms per dwelling? Comment on the suburbs that average more than eight rooms per dwelling.

### Add your reply and code here:

